---
title: "Class 5: Data Viz with"
author: "Mudit"
format: gfm
---
R has a lot of ways to makefigures and graphs in particular.
One that comes with R out of the box is called **"base" R** - the `plot()` function

```{r}
plot(cars)
```

A very popular package in this area is called **ggplot2**

Before I can use an add-on package like this, I must install it with `install.packages("ggplot2")` command/function.

Then to use the package, I need to load it with a 'library(ggplot2)' call.

```{r}
library(ggplot2)
ggplot(cars) + 
 aes(x=speed, y = dist) +
  geom_point()
```

For "simple" plots like this one base R code will be much shorter than ggplot code.

Let's fit a model and show it on my plot

```{r}
ggplot(cars) + 
 aes(x=speed, y = dist) +
  geom_point() + 
  geom_smooth()
```

ggplot work like a layer where each '+' adds a layer

Every ggplot has at least 3 layers

- **data** (data.frame with the numbers  and stuff you want to plot)
- **aes**thetics (mapping of your data columns to your plot)
- **geom**s (there are tones of these, basics are `geom_point()`, `geom_line()`, `geom_col()`)

```{r}
head(mtcars)
```

Make me a ggplot of the `mtcars` data set using `mpg` vs `disp`

```{r}
ggplot(mtcars) + 
 aes(x=mpg, y = disp) +
  geom_point() + 
  geom_smooth()
```

Make me a ggplot of the `mtcars` data set using `mpg` vs `disp` 
and set the size of the points to the `hp`
and et the color to `am`

```{r}

ggplot(mtcars) + 
 aes(x=mpg, y = disp) +
  geom_point(aes(size = hp, col = am)) + 
  geom_smooth()
```
```
Now colour all points blue

```{r}
```{r}

ggplot(mtcars) + 
 aes(x=mpg, y = disp) +
  geom_point(aes(size = hp), col = "blue") + 
  geom_smooth()

```
```

```{r}
```{r}
library(ggrepel)
ggplot(mtcars) + 
 aes(x=mpg, y = disp, size = hp, col = am, label = rownames(mtcars)) +
  geom_point() + 
  geom_smooth() + 
  facet_wrap(~am) +
  geom_text_repel()
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)

p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()

p + scale_colour_manual( values=c("blue","gray","red") ) +
  labs(x = "Control (No Drug)", y = "Drug Treatment")

ggsave("myplot.pdf")
```



```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)

library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)

ggplot(gapminder_2007) +
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop) +
  geom_point(alpha=0.5)

ggplot(gapminder_2007) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop), alpha=0.5) + 
  scale_size_area(max_size = 10)
```

